---
title: "Think like a programmer - an introduction"
date: "`r format(Sys.time(), '%A %d, %B')`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
    theme: readable
    colortheme: "default"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, results="hide", fig.show="show")
```

# Statements and loops

Statements and loops are another cornerstone of programming. Statements and loops are of different kinds and allow you to perform operations when some conditions are, or are not, met.

You will use a statement or a loop when you will:

 - need to decide whether to watch a movie or not;
 - be eating ice cream until the box is empty;
 - be knitting winter socks for each friend you care about.
 
And yes, all of these are very nice and meaningful activities, especially knitting socks for your friends.

## If-else statement

The first loop we will talk about...is not actually a loop. It's more a test for one, or more, conditions. It is called _statement_.

The `if` and `if-else` statement is used to evaluate if a condition, or more, is met. If yes, or if not, then something will happen (or not!).

An example of `if-else` statement will be used to determine what to do tonight:

"If I have popcorn and zombie movie, I will watch the movie".

First, you need to translate this sentence into something that resemble code, i.e. this is called _pseudocode_:

```{r eval = FALSE}

if (popcorn & zombie_movie) {
  print("I will watch the movie")
}

```

As you can see, an `if` statement has a specific structure:
  
  - a keyword defining the kind of statement/loop, in this case `if-else`;
  - a set of parenthesis `()` which contains the condition that is evaluated;
  - another set of parenthesis `{}` which contains the instruction that will be performed;
  - indentation: whatever is in between the `{}` is usually shifted to the right (with **Tab** button).

Every statement/loop will have these elements. Indentation is not necessary but it's useful to keep the code readable: every time you open the  `{}`, shift to the right one tab.

Now, let's add add a further specification to your previous attempt.

The new specification will define what is happening if "I can't have popcorn and zombie movie":

"If I have popcorn and zombie movie, I **will not** go to bed".

Let's write some more pseudocode:

```{r eval = FALSE}

if (popcorn & zombie_movie) {
  print("I will watch the movie")
} else {
  print("I will go to bed")
}

```

An `if-else` statement can be used in its `if` form, in the `if-else` form, or in more complicated forms that adds further layers to the `else` part. In all cases, R will evaluate the first `if`. If the condition is not satisfied, R will evaluate the `else` or the `elseif` and so on.

```{r eval = FALSE}

if (popcorn & zombie_movie) {
  print("I will watch the movie")
} else if (zombie_movie) {
  print("I may watch the movie")
} else {
  print("I will go to bed!")
}

```
 
It is possible that none of the conditions are satisfied. In this case nothing of what was placed between `{}` is going to happen.
 
Now, let's define the variables `popcorn` and `zombie_movie` as logical values:

```{r eval = FALSE}

# define variables as logical values
popcorn <- TRUE
zombie_movie <- FALSE

# test the variables
if (popcorn & zombie_movie) {
  print("it's going to be a fun evening!")
} else if (zombie_movie) {
  print("not everything is lost!")
} else if (popcorn) {
  print("at least something to eat!")
} else {
  print("i'll go to bed!")
}

```

Of course, if you change the values of `popcorn` and `zombie_movies`, the result may change.

In R, there is also a compact version of the `if-else` statement: the `ifelse()` function. It works the same way as a loop:

```{r}

# define variables as logical values
popcorn <- TRUE
zombie_movie <- FALSE

# let's define the ifelse() function
ifelse(popcorn & zombie_movie, "i'll watch the movie", "i'll go to bed!")

```

The idea, here, is that the `ifelse()` function takes three parameters as input:

 1. condition to evaluate;
 2. what happens if condition is TRUE;
 3. what happens otherwise, i.e. if condition is FALSE.

## For loop

A `for` loop is an actual loop in the sense that it does something for a fixed amount of times.

In other words, a `for` loop is used to perform the same operation many times over a set of objects. As we mentioned in the example: "Knitting winter socks for each friend you care about", in a `for` loop way of thinking, becomes something like:

```{r eval=FALSE}

for (friends_i_care_about) {
  knit a pair of (nice and warm) socks for a friend
}

```

This is how to translate "Knit winter socks for each friend you care about", into something that resemble a `for` loop. Now, let's make actual code that creates warm and nice socks for each one of your friends:

```{r results="hold"}

# define some friends
friends <- c("Francesca", "Valerio", "Sofia", "Elena")

# knit a personalised pair of socks for each one of them
for (a_friend in friends) {
  # knit and print the socks!
  print(paste0("knitting nice and warm socks for ", a_friend, "!"))
  cat("   |---|       |---|\n")
  cat("   | ", substr(a_friend, 1, 1), " |", "       | ", substr(a_friend, 1, 1), " |\n", sep="")
  cat("   |---|___    |---|___\n")
  cat("   |______/)   |______/)\n\n")
}

```

As you can see, the `for` loop allowed you to knit a pair of socks for each one of your friends, which is very nice.

What is even nicer is that you can use the `for` loop for many more purposes but the basic idea is always the same: whenever you need to perform some repetitive operations over a set of objects, i.e. numbers, lists, columns in a dataframe, then the `for` loop is the tool you need.

## While loop

A `while` loop is another actual loop, in the sense it does something for a fixed number of times.

However, contrary to a `for` loop, the `while` loop works _until_ a condition is, or is not, satisfied: "I will eat ice cream until the box is empty". Once the box is empty, it's time for you to either 1) open another box, or 2) stop eating ice cream. Both ways, the condition _until the box is empty_ was satisfied and the loop is finished.

Let's make a nice visual example with ice cream:

```{r results="hold"}

# how many scoops of ice cream do I have
remaining_scoops <- 10

# while I have remaining scoops available let's eat them
while (remaining_scoops != 0) {
  # the scoops
  all_the_scoops <- "OOOOOOOOOO"
  # the box
  empty_box <- "__________"
  # the flavours
  flavours <- c("dark chocolate", "hazelnut", "banana", "coffee, mmmmm", "mango, why is mango here??", "vanilla", "strawberry??? oh, no!", "coconut", "dark chocolate, again, yes!", "mint, and...")
  cat("|", substring(all_the_scoops, 1, remaining_scoops), substring(empty_box, 1, 10-remaining_scoops), "|   ", flavours[11-remaining_scoops], "\n", sep="")
  # a scoop is gone
  remaining_scoops <- remaining_scoops - 1
}
cat("|__________|   oh oh!")

```

## Combining loops

So far we made it through two actual loops (`for` and `while`) and a statement (`if-else`) and it is now that the real fun begins. As you already guessed, we can combine all the loops and all the statements you might think of, in order to get some code doing some very wild things.

Have a look at the file `log_example.R` for some ideas :-)

## Some very nice exercises

### If-else

Create a `if-else` loop that tests whether a person is in your contact list and prints its phone if the person is found. Of course, you will need a contact list, first: load the file `contact_list.csv`. The field separator is tab, i.e. `\t`.

 - Idea: `length` and `which()`;
 - Idea: what is `%in%` doing?

Can you do the same in one line using the function `ifelse()`?

### For and While

When we discussed how to write a function to compute the average of a set of values, we started doing it for vectors of two, and then we generalised to any vector.

To generalise, we used the function `sum()`.

Can you replicate what the function `sum()` is doing:
 
 - using a `for` loop?
 - using a `while` loop?

Can you use the loop to re-write the `arithmetic_mean()` function?
