---
title: "Think like a programmer - an introduction"
date: "`r format(Sys.time(), '%A %d, %B')`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
    theme: readable
    colortheme: "default"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, results="hide", fig.show="show")
```

# Short intro to R

R is a programming language which is mainly used for statistical purposes but in reality it has many more applications. R exists in a base form, offering a wide set of tools, which are called _functions_, that are used to perform many operations such as computing means, plotting, regression analysis, and much more.

In addition, there are many packages, e.g. _ggplot2_, _phyloseq_, _Vegan_, ..., that were created to extend the basic set of R functions for more specific tasks like DNA sequences analysis, gene enrichment analysis, network construction, advanced plotting, computing $\alpha$-diversity, and much more.

Of course, packages need to be installed before using them. You can find packages [in several places](https://rdrr.io/find/?repos=cran%2Cbioc%2Crforge%2Cgithub&fuzzy_slug=), such as CRAN, R-Forge, Github, or Bioconductor.

# Super short into to RStudio

RStudio is and Integrated Development Environment (IDE) that allows you to run R scripts, edit them in a text editor, visualise and export results and plots, look into tables, create and compile Markdown-based documents (such as this document we're using), and many more things.

If you don't want to use RStudio you can use R from the command line and write your scripts using a text editor.

# Think like a programmer

Programming may seems complicated at first, and R is not the best way of learning it (there are programming languages that are easier to understand), but it's not about the language. Programming is a different way of thinking how problems can be solved.

In programming you should think through logic statements and loops.

Programming means putting together a set of instructions, in a _program_ or a _script_, that will return the expected result. These instructions _have to_:

-   be ordered, since it is not possible to read a book with random sentences;
-   follow that-programming-language specific rules.

These instructions _should be_:

-   as simple as possible;
-   well commented so (you and) other peoples can understand it in the future.

# R environment

When you first start a session of R, you must be aware that R is _sitting_ in a specific folder in our computer. You can, of course, specify which folder this folder is, but you need to be aware. So...which folder are you currently in, i.e. the working directory? You can ask R using the function `getwd()` which basically _gets the working directory_:

```{r}

# where am I??
getwd()

```

If this is not the directory you want to be working in, and there are many reasons for that, you can always change it using another function which is `setwd()`, which basically _sets a Working Directory_:

```{r}

# move to the folder containing the data for this course
setwd("/home/gabriele/work/microbiology/think_like_a_programmer/")

```

Now that you set the working directory, it's time to get to know the R environment, which represent the sets of variables, functions, and packages that are available to you in a specific R session. To get these information you can use two functions: `ls()` and `search()`:

```{r}

# show objects I can use
ls()

# show packages I can use
search()

```

The working directory, and the working environment are very important.

You can also have an idea of which files are currently in the working directory, i.e. the one that `getwd()` returns, using the function `list.files()`:

```{r}

# show the files in the current directory
list.files()

```

# Variables

A variable can be considered as a box that contains zero, one, or many things which can be used, modified, and deleted whenever you like.

You have probably heard about variables. Variables are used, for example, in mathematics to represent unknown values in equations:

$3x + \frac{1}{2} y = 0$

In this case $x$, $y$ are the names of two different variables. They can take, _theoretically_, any value and you will get something back. The same applies in R: a variable can be _substituted_, or can _contain_ any value, and is used to perform tasks.

When you use, or _call_, a variable, what you are actually doing is using the value that is contained in it.

So, a variable is like a box which can contain something that you want to use later on. That is why you put it into the box: it's just a storage. When you need whatever was inside the box, you will know where to find it and how it's called.

The action is performed _on_ the content of the box. Of course, if something is not needed later on, you don't define a variable for it: it's a waste of space (how many things you have in boxes that are never used and just takes space??).

Assigning a value to a variable in R:

```{r}

# assign a value using left arrow
a_value <- 5
a_value

# assign a value using right arrow
10 -> another_value
another_value

# assign a value using equal
a_third_value = 15
a_third_value

# assign two values at once
left_assign <- 100 -> right_assign
left_assign
right_assign

```

All of these assignments are valid. However, the symbol `=` is usually found only when a value is _passed_ to a parameter, while the left arrow, i.e. `<-`, is the preferred one. The right arrow can also be used but it's very uncommon.

# Data types

In variables you can store objects. These objects can be classified in data types according to their nature.

In R there are several data types, each one with its own characteristics. A general distinction can be made by separating data types in _numerical_, _categorical_, and _logicals_.

## Numerics

Numerics, as you may expect, are numbers. Numbers which can be used to perform mathematics (, statistics, and much more). Numerics can be either integers or double (if they have decimals).

Example of numerics:

```{r}

# define a variable named one and assign an integer to it
one <- 1

# define a variable named three_point_five and assign a double
three_point_five <- 3.5

# sum up one and three_point_five, you will get a double
one + three_point_five

# same result can be obtained by summing up 1 and 3.5
1 + 3.5

```

Numerics can be stored inside variables and then used for operations, at any point in the code, but only after they were _declared_, i.e. after they actually exist.

The only thing you need to remember is the name of the variable where the numeric was stored. Also, a good programmer knows that there are many golden rules. One of them is:

> Name variables in a way that helps you remember what they contain.

Which will help in keeping the code as simple as possible.

You can now use the variables that were defined above, to perform some more operations:

```{r}

# some more random operations
one + 3.5

three_point_five / 3

one * three_point_five

```

## Categorical - characters

Characters, on the other hand, represents those _symbols_ that are not numbers. So, for example, letters and words.

Example of characters:

```{r}

# define a variable named a_characters containing a single character
a_character <- "c"

# define a variable where many characters are pasted together
# this sequence of characters is called "string"
some_characters <- "manycharacters"

```

Clearly, arithmetic operations can be only performed on numbers, i.e. numerics. It does not make sense to sum up characters, or does it?

```{r error=TRUE}

# sum (the content of) some_characters and 1
some_characters + 1

```

Here, R is telling you that something is wrong. The errors arose because you are trying to use a non-numeric object, i.e. the variables `some_characters` to perform some arithmetic operations (a sum).

The interesting thing with `characters`, is that they can be concatenated to form more complex sequences of `characters`. To do this, you will need a new function: `paste()` that allows you _combining_ the content of variables together.

For example, define a path to a folder in your computer:

```{r}

# path to Desktop folder
path_to_folder <- "/home/gabriele/Desktop/"

# show the path
path_to_folder

# path to work-related stuff. put together path_to_folder with the
# name of the work folder:
path_to_work_folder <- paste(path_to_folder, "work_folder/", sep="")

# show the path
path_to_work_folder

# path to pics from the last vacation. same as for the work folder
path_to_vacations_pics_folder <- paste(path_to_folder, "vacations_pics/", sep="")

# show the path
path_to_vacations_pics_folder

```

In this way you can define a common path, `path_to_folder`, and build more complex paths on top of it. That's how variables can be pasted:

```{r}

# set result in advance
five <- 5

# now sum three + two
print(paste("three plus two equals", five))
```

### A very nice exercise

Use the function `list.files()` to view the content of _path_to_folder_ and _path_to_work_folder_. Then, move the current working directory (remember `getwd()`) to _path_to_folder_.

## Categorical - factors

Factors may look like characters, but they have different properties.

Examples of factors:

```{r error=TRUE}

# these are characters
some_characters <- "some_characters"

# define a factor using a sequence of characters
a_factor <- as.factor(some_characters)

```

Factors are used to represent _Categorical variables_ (in the statistical sense). _Categorical variables_ can be _Nominal_ or _Ordinal_. The difference is that _ordinal_ have a sort of order, _nominal_ do not.

For instance, define a nominal categorical variable that contains many different (sequences of) characters, such as:

```{r}

# list some friends as characters
friends <- c("Francesca", "Valerio", "Sofia", "Elena")

# cast friends to factor
friends <- factor(friends)

# show the content of the variable
friends

# add other friends, while keeping same levels
friends <- c("Francesca", "Valerio", "Sofia", "Elena", "Francesca", "Valerio")
friends <- factor(friends)
friends

```

Now, define an ordinal categorical variable, instead:

```{r}

# list soil depths, as characters
depths <- c("surface", "medium", "deep", "shallow", "very_deep")

# cast depths as factor and order it using levels
depths <- factor(depths, order=TRUE, levels=c("very_deep", "deep", "medium", "shallow", "surface"))

# show the content of the variable
depths

```

There is an interesting difference between `friends` and `depths`. The first one has levels that are ordered, by R, alphabetically: they are names, meaning no one is bigger, slower, higher, or weigh more than another: just names. And yes, you can still order them as you like, but you need to specify it as with the `depths`, i.e. using `levels`.

On the other hand, soil levels are ordered since _verydeep_ is deeper than _medium_ which, in turn, is deeper than _shallow_. Here, the order is important. That's why you may want to set the levels (this is very handy when performing statistical analysis or other stuff such as plotting).

## Logicals

Logical can take values that are either `TRUE` or `FALSE`. They are fundamental in programming because contrary to human logic, computer logic works with zero and one, black or white: in programming a _statement_ can be `TRUE` or `FALSE`.

Logical values have specific operations that are performed through logical operators. They are three: AND, OR, and NOT. In R, they are written as symbols which you can find on your keyboard:

-   AND: `&`;
-   OR: `|`;
-   NOT: `!`.

`TRUE` can be abbreviated as `T`, `FALSE` as `F`.

The AND operator takes two logical values as input, and return TRUE when both are TRUE. This can be translated as: "I will be happy if I can have popcorn and a zombie movie!".

```{r}

# true and true
TRUE & TRUE

# true and false
TRUE & F

# false and true
FALSE & T

# false and false
F & F

```

The OR operator works in the same way as the AND operator but it returns TRUE if one of the two logical values is TRUE. This can be translated as: "I will be happy if I can have either a hot chocolate or a coffee with whip cream".

```{r}

# true or true
T | T

# true or false
TRUE | FALSE

# false or true
F | TRUE

# false or false
F | FALSE

```

The NOT operator is a negation of the logical operator it is used with logical values or functions:

```{r}

# not true
!T

# not false
!F

# is 4 a numeric??
four <- "four"
is.numeric(four)

# or, is it not??
!is.numeric(four)

# wait, but is it or is it not??
!!is.numeric(four)

```

There are other operators that are used for comparing objects: 

 - equal to: `==`;
 - not equal to: `!=`;
 - greater than: `>`;
 - greater than or equal to: `>=`;
 - smaller than: `<`;
 - smaller than or equal to: `<=`.

A few examples:

```{r}

# equal to
"hello" == "ciao"

# equal to
4 == 5

# not equal to
10 != 100

# smaller than
4 < 5

# greater than or equal to
100 >= 10

# you can also combine expressions
10 > 9 | "a" == "b"

10 < 9 & 5 > 4

10 >= 9 | 5 < 4

```

Remember that logical operators have precedence, for example as in mathematics when multiplication has precedence over addition. For R:

 - `!` has precedence over
 - `&` which has precedence over
 - `|`.
 
For a full list of operators and precedence, see [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html).

### Vectors of logical values

Logical values can be combined in vectors:

```{r}

# define two vectors of logical values
logi_vals_one <- c(TRUE, TRUE, FALSE)

logi_vals_two <- c(FALSE, TRUE, FALSE)

# test the AND logical operator on the vectors
logi_vals_one & logi_vals_two

# test the OR logical operator on the vectors
logi_vals_one | logi_vals_two

# test the NOT logical operator on the vectors
!logi_vals_one
!logi_vals_two

```

### Side note about Casting

To _cast_ (or _casting_) means transforming an object of a type into an object of another type, as you just did using the function `as.factor()`. Another example can be done using the function `as.integer()`. This function transforms something into an integer, if it is possible:

```{r}

# transform a double into integer
as.integer(2.3)

# transform a character into an integer and sum two
as.integer("1") + 2

# transform another character into an integer
# (this will give you an error since the letter A does not correspond to any number!)
as.integer("A")

```

Interestingly, you can perform operations using factors:

```{r}

# sum up some depths
sum(as.integer(depths))

```
