---
title: "Think like a programmer - an introduction"
date: "`r format(Sys.time(), '%A %d, %B')`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
    theme: readable
    colortheme: "default"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, results="hide", fig.show="show")
```

# Data structures

A data structure is a container for a set of objects which can be of many type, e.g. numerics, characters, factors, or logicals.

Data structures are usually stored into variables, exactly like a number, or a sequence of character.

There are different type of data structures, in R.

## Vectors

Vectors are variables containing more than one object. You have already seen a vector:
  
```{r}

# define a character variable, which is actually a vector
friends <- c("Francesca", "Valerio", "Sofia", "Elena")

# print the content of friends
paste("the content of friends is ", friends, sep="")

```

`friends` is a vector, i.e. a data structure. The vector `friends` was built using the function `c()` that basically combines the characters that were passed as input. In `friends` four values were stored. You can get this information using a function that show you the _length_, or the number of elements, in a vector. The function, as you may have guessed, is `length()`.

```{r}

# how many elements are there in friends?
length(friends)

# print the objects contained in the vector called friends
print(paste0("the first object in contact is ", friends[1]))
print(paste0("the second object in contact is ", friends[2]))
print(paste0("the third object in contact is ", friends[3]))
print(paste0("the fourth object in contact is ", friends[4]))

```

Vectors can also be filled, after they were defined:

```{r}

# define an empty vector
empty_vector <- vector()

# put something in position 1
empty_vector[1] <- "position_one"

# put something in position 10
empty_vector[10] <- "position_ten"

```

It is important to note that, in R, the first value of a vector starts at position 1. In other programming languages, this is not always true since some may start at position 0.

#### Question

Can you create a variable with one value and access this value as you would access a value in a vector?

## Lists

Lists are data structures containing more than one object, as vectors. They are defined in the same way:
  
```{r}

# define an empty list
a_list <- list()

# add something to the list, i.e. a numerical value
a_list[["a_number"]] <- 121.36509356

# add something more to the list, i.e. some characters
a_list[["some_chars"]] <- "somechars"

# print the content of the list
a_list

# define a list and fill it in one line
one_line_list <- list(a_number=4, a_char="a", some_logicals=c(T,F,T,T))

# print the content of the one line list
one_line_list

```

As you noticed, lists can contain different data types at the same time **but not in the same element**:

```{r}

# define empty list
a_complex_list <- list()

# an element of numerics is ok
a_complex_list[["numerics"]] <- c(1, 1.24587, 45, 39372.2134)

# an element of characters is ok
a_complex_list[["friends"]] <- c("Francesca", "Valerio", "Sofia", "Elena")

# an element of numerics AND characters is not ok!
a_complex_list[["nums_and_chars"]] <- c("one", 2, "three", 4, 5)

# print the content of the vectors we just created
a_complex_list[["numerics"]]
a_complex_list[["friends"]]
a_complex_list[["nums_and_chars"]]

```

If we look at the `nums_and_char` element, you will notice that numerics are quoted. This means that they are not numerics any more. They were _casted_, automatically by R, into characters. This happened because R can't cast a character into a numeric so that's the only way to keep data types consistent in the element:
  
```{r}

# print the type of data contained in all the elements of the list
print(c(typeof(a_complex_list[["numerics"]]), typeof(a_complex_list[["friends"]]), typeof(a_complex_list[["nums_and_chars"]])))

```

## Matrices

Matrices are collections of vectors, which means the can contain only one data type:
  
```{r}

# define a matrix of zeroes
a_matrix_of_numbers <- matrix(0, nrow=3, ncol=4)

# define a matrix of chars
a_matrix_of_chars <- matrix("0", nrow=3, ncol=4)

# show the variables content
a_matrix_of_numbers
a_matrix_of_chars

# assign a character to a_matrix_of_numbers
a_matrix_of_numbers[1, 1] <- "0"

# show the content of the variable
a_matrix_of_numbers

```

After substituting an element of `a_matrix_of_numbers` with a character, i.e. "0", all the other zeros were _casted_ into chars.

With this example, an element of the matrix. was _substituted_ with a character. This means that you can interact with the matrix itself. Substitution, or _assignment_:
  
```{r}

# set value in position (1, 1)
a_matrix_of_numbers[1, 1] <- "0"

```

By doing so, you are accessing the value in the matrix, i.e. row 1, column 1, and putting something _into_ it, or _assigning_ something to that position. You _substitute_ or _overwrite_ the previous content.

Of course, you can access whole matrices, a single cell, a whole row, or a whole column. The general notation follows this pattern `matrix[n_of_row, n_of_column]`:
  
 - if we specify `n_of_row` **and not** `n_of_column`, i.e. `a_matrix[1, ]`, we are accessing the first row, all the columns. We are getting the whole first row;
 - if we specify `n_of_column` **and not** `n_of_row`, i.e. `a_matrix[, 1]`, we are accessing the first column, all the rows. We are getting the whole first column;
 - if we specify both `n_of_row` **and** `n_of_column`, i.e. `a_matrix[1, 1]`, we are accessing a single cell;
 - if we specify both `n_of_row` **and** `n_of_column` but more than one, i.e. `a_matrix[c(1,2), c(3, 4)]`, we are accessing a subset of cells;
 - if we specify nothing, i.e. `a_matrix[,]` **or** `a_matrix`, we are accessing the whole matrix.

An example with numbers:
  
```{r}

# define a matrix with 3 cols and 4 rows, with numbers from 1 to 12
# (we have 12 cells, right? and R will fill the matrix for us)
a_matrix_of_numbers <- matrix(c(1:12), ncol=3, nrow=4)

# print
a_matrix_of_numbers

# now, grab the second row, all the columns
a_matrix_of_numbers[2, ]

# now, grab the second column, all the rows
a_matrix_of_numbers[, 2]

# now, grab the second row, third column
a_matrix_of_numbers[2, 3]

# now, grab the second and fourth rows, the second and third column
a_matrix_of_numbers[c(2, 4), c(2, 3)]

```

Pretty convenient to subset matrices and work with some of the values, if needed!
  
What is even more convenient, however, is the possibility to assign names to rows and columns:
  
```{r}

# define a matrix with 3 cols and 4 rows, with numbers from 1 to 12
a_matrix_of_numbers <- matrix(c(1:12), ncol=3, nrow=4)

# use the function rownames to assign names to rows
rownames(a_matrix_of_numbers) <- c("row_one", "row_two", "row_three", "row_four")

# use the function colnames to assign names to columns
colnames(a_matrix_of_numbers) <- c("col_one", "col_two", "col_three")

# remember that the function rownames (and also colnames)
# can be used to visualise columns and rows names
rownames(a_matrix_of_numbers)
colnames(a_matrix_of_numbers)

```

`colnames()` and `rownames()` are two-way functions: they can show you something or you can use them to assign something.

At this point, you can get columns, or rows, by using their names:
  
```{r}

# get second column
a_matrix_of_numbers[, "col_two"]

# get fourth row
a_matrix_of_numbers["row_four",]

```

Matrices can be combined together, by columns or by rows, using two functions: `cbind()` and `rbind()`:
  
```{r error=TRUE}

# define a matrix with 3 cols and 4 rows, with numbers from 1 to 12
matrix_one <- matrix(c(1:12), ncol=3, nrow=4)

# define another matrix with 3 cols and 4 rows, with numbers from 1 to 12
matrix_two <- matrix(c(13:24), ncol=3, nrow=4)

# define a matrix with 4 cols and 3 rows, with numbers from 1 to 12
matrix_three <- matrix(c(25:36), ncol=4, nrow=3)

# now merge matrix_one and matrix_two by columns, i.e. use cbind
cbind(matrix_one, matrix_two)

# now merge matrix_one and matrix_two by rows, i.e. use rbind
rbind(matrix_one, matrix_two)

# now try to merge matrix_two and matrix_three by rows, i.e. use rbind
rbind(matrix_two, matrix_three)

```

The last `rbind()` did not work because the matrices has a different number of columns. When putting together matrices, you need to make sure they have the correct dimensions.

### How bugs are born

In linear algebra matrices are quite popular. So, it is important to know how to use them to perform common operations such as sum or multiplication, since it may be non obvious. An example of multiplication:
  
```{r}

# define a simple 2x2 matrix
simple_matrix <- matrix(c(1,2,3,4), nrow=2, ncol=2)

# define simple vector to multiply the simple matrix
simple_vector <- c(0, 1)

# multiply matrix by vector, "R version"
simple_matrix * simple_vector

# multiply matrix by vector, linear algebra version
simple_matrix %*% simple_vector

```

This is a good example of how, using the wrong tool, i.e. `*`, to perform a simple operation will produce what is called a _bug_. This is an error that give you unexpected, or wrong results.

Some bugs are easy to spot because your code won't work. Some others are more complex, because you might not notice them. Be careful!

## Data frames

Data frames, contrary to matrices, can contain any data type:
  
```{r}

# create a dataframe with 4 rows and 2 columns
a_dataframe <- data.frame(some_numbers=c(1, 435, 65432, 7.65376), some_chars=c("Elena", "Sofia", "Valerio", "Francesca"), some_logicals=c(T, F, F, F))

# show the content of the dataframe
a_dataframe

```

You just defined a dataframe and assigned a name to each column. This was done by telling to R that `some_numbers`, `some_chars`, and `some_logicals`, i.e. three vectors, should be combined into a `data.frame`. The process in all its parts is:
  
```{r}

# first, define two vectors
student_mark <- c(10, 8, 8.5, 9)
student_name <- c("Elena", "Sofia", "Valerio", "Francesca")

# second, put the two vectors together as a data frame
students_data <- cbind.data.frame(student_mark, student_name)

# show the content of the dataframe
students_data

```

Here, the function `cbind.data.frame()` was used to put together two vectors as columns. You could do it by row, i.e. using `rbind.data.frame()`. It's exactly the same as using `rbind` or `cbind` for matrices.

Data frames, as matrices, can be accessed in several ways. You can access a single cell, a column, a row, or a set of columns and rows by using the same notation, i.e. the square brackets `[1, c(1, 4)]`, but also using row and columns names. In the case of columns, though, it's a bit different with respect to what you can do with matrices:
  
```{r}

# access the numbers column using the $ symbol
students_data$student_mark

# access the numbers column using the $ symbol
students_data$student_name

```

Remember that data frames can be accessed in all the way matrices can be accessed, but they also have the `$` symbol allowing you to get a column by using its name (we can't use the `$` with matrices). The `$` can be used to add new columns to data frames in a very simple way:

```{r}

# add new column, as a vector, to an existing dataframe
# by specifying its name
students_data$student_surname <- c("F.", "G.", "M.", "B.")

# show the content of the new dataframe
students_data

```

### Last, small thing about lists

As a side note about lists, it is worth saying that you can store anything, in a list:

```{r}

# create an empty list
another_complex_list <- list()

# add a numeric vector to the list
another_complex_list[["numerics"]] <- c(1, 2, 3, 4, 5)

# add a character vector to the list
another_complex_list[["chars"]] <- c("1", "2", "3", "4", "5")

# add a matrix to the list
another_complex_list[["matrix"]] <- a_matrix_of_numbers

# add a dataframe to the list
another_complex_list[["dataframe"]] <- students_data

# show the content of the list
another_complex_list

```

In this example, a list was used to create a complex data structure containing many different piece of information. Each piece of information is stored into it's own cell, with it's own name, to easily access it when needed.

### A side note about tables

Matrices and data frames can be build from existing tables, that you may have built outside R such as the data you managed to collect after sampling soil, extracting DNA, sequencing it, and everything in between.

The only thing that you need to know about tables is:

 > when you build a table make sure it is well formatted and save it as _.csv_.

If this rule is applied your (and other people's) life will be easier and happier.

To load a (well formatted) table from a file you will need to know where it is located in your computer:

```{r}

# set path to working directory
path_to_working_directory <- "/home/gabriele/work/microbiology/"

# set path to directory where files are kept
path_to_files <- "think_like_a_programmer/R_foundations/lessons_files/data/"

# load a table from a csv file
ecology_data <- read.csv(paste(path_to_working_directory, path_to_files, "meta_data.csv", sep=""), header=T, sep="\t")

# show number of rows and number of columns
nrow(ecology_data)
ncol(ecology_data)

# show these values at once
dim(ecology_data)

# show the first six lines of the loaded table
head(ecology_data)

# show the first ten, but you need to specify it
# since by default the lines that are shown are six
head(ecology_data, n=8)

```

At this point, it's worth introducing three functions that can be used to subset tables. These function take advantage of logical values. In other words, these functions will give you back some values will be `TRUE` or `FALSE`.

To subset a table you will need to define _what you like_, which will be `TRUE`, and _what you don't like_, which will be `FALSE`.

Think like a programmer, now. First, you must know what you're dealing with, in this case a table. Have a look at the table:

```{r}

# print table, just to have an idea of what we are dealing with
head(ecology_data)

```

First: you will need to work with the column Treatment, i.e. _ecology_data$treatment_.

At this point, you need to get those samples which treatment is equal to _Control_. You can use logical operators:

```{r}

# get control samples
ecology_data$treatment=="C"

```

You can store these vector of logical values inside a variable or you can directly use the logical values:

```{r}

# get control samples
treated_samples <- ecology_data$treatment=="C"

# show subsetted table using the variable just created
ecology_data[treated_samples, ]

# show subsetted table using logical values directly
ecology_data[ecology_data$treatment=="C", ]

```

There is an interesting function, that you can also use for other purposes. The function is `which()`:

```{r}

# get control samples using the which function and a logical operator
which(ecology_data$treatment=="C")

```

#### Question

What is `which()` telling you?

The `which()` function returns the number of rows of those samples that satisfied the condition. Using the logical operator `==` without the `which()` only returns a vector of logicals that tells you which samples are controls, i.e. `control==TRUE`, and which are not, i.e. `control==FALSE`.

So, now that you know which samples are control and which are not, it's time to use this piece of information to subset the original metadata table:
  
```{r}

# store control samples in a variable
control_samples_rows <- which(ecology_data$treatment=="C")

# subset ecology_data, by rows, since the which returned
# the number of rows for control samples
ecology_data[control_samples_rows, ]

# or, use the logical vector
control_samples_logical <- ecology_data$treatment=="C"

# subset
ecology_data[control_samples_logical, ]

```

Of course, it is possible to create any combination of conditions, using any (combination of) logical operator.

#### Two very nice exercises

Subset the table to get only those samples that:
  
  1. are not Controls;
  2. are Controls sampled in Winter.
