---
title: "Think like a programmer - an introduction"
date: "`r format(Sys.time(), '%A %d, %B')`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
    theme: readable
    colortheme: "default"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, results="hide", fig.show="show")
```

# Plotting

# Base Plotting

This is what you need to know about plotting:

> Plotting is always tricky but one thing is for sure: do not use pie-charts

For all the other best practices about plotting, [this is the website](https://www.data-to-viz.com/caveats.html) that you want to check out.

## A very basic scatterplot

Plotting in R doesn't change much between the type of plot. Either a scatterplot, a lineplot, an histogram...once you get the idea of how it works, then it's done (although, I still have to grasp this idea myself).

Let's plot some data, then:

```{r}

# load a table from a csv file
a_nice_table <- read.csv("/home/gabriele/work/microbiology/think_like_a_programmer/R_foundations/lessons_files/data/a_nicely_formatted_table.csv", header=T, sep=",")

# let's use Sepal length and width to plot a simple scatterplot
plot(x=a_nice_table$Sepal.Length, y=a_nice_table$Sepal.Width)

# let's plot a histogram, now
hist(a_nice_table$Sepal.Width)

# and a boxplot
boxplot(Sepal.Length ~ Species, data = a_nice_table)
```

Not bad, but you can do better. For instance, add some color, why not? Also, the axis labels don't look very nice. What can you do?? Let's look at what the R manual has to say regarding the `plot` function.

```{r}
help(plot)
```

Ok, but how to assign colours to data points? You may want to colour the data points with respect to the species they belong to.

To do it, you will need a colour for each species, i.e. each category. It's now time to think like the programmer you are. Some questions you should ask yourself:
 
 1. How do I get a set of colours?
 2. How many colours do I need?
 3. How do I even create a set of colours?
 4. How big should this set be?
 
Let's try to answer:

 1. [R colours, here](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf)
 2. how about one color `for` each category/species?
 3. what about a vector?
 4. it's definitely possible that I will need a colour for each data point, right?
 5. or even a shape?? for shapes, see [here](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/points)
 
If you think these answers make sense...it's time to worry about how to assign a colour to a category, i.e. how to actually make answer n. 3 a reality? Again, let's think programming. What we know:

 1. the number of data points;
 2. the category of each data point, i.e. the species;
 3. which colours to use;
 4. a vector is needed and it must be as long as the number in `1.`;
 5. in answer n. 2 (see above) there is a very specific keyword.
 
Side note: shape, for R are called **pch** (probable meaning of **pch** is _point character_). So if you want to know which shapes are available, now you have the keyword.

Always remember that, if you want to fill the points you see in the plot, you need to set the shape using one **pch** that allows you to have a _fill_.

Now, using the ideas above, plot:

```{r}

# colours I like
colours_i_like <- c("lightcoral", "midnightblue", "saddlebrown")

# shapes I like
shapes_i_like <- c(21, 23, 25)

# empty vector with a colour for each data point
data_colours <- vector()

# empty vector with a different shape for each data point
data_shapes <- vector()

# for each category/species
for (sp in a_nice_table$Species) {
  # get its colour and shape
  if (sp=="setosa") {
    # assign colour
    data_colours <- append(data_colours, colours_i_like[1])
    # assign shape
    data_shapes <- append(data_shapes, shapes_i_like[1])
  } else if (sp=="versicolor") {
    # assign colour
    data_colours <- append(data_colours, colours_i_like[2])
    # assign shape
    data_shapes <- append(data_shapes, shapes_i_like[2])
  } else {
    # assign colour
    data_colours <- append(data_colours, colours_i_like[3])
    # assign shape
    data_shapes <- append(data_shapes, shapes_i_like[3])
  } 
}

# have a look at data colours and shapes
data_colours
data_shapes

# use data_colours to color the data points
plot(x=a_nice_table$Sepal.Length, y=a_nice_table$Sepal.Width, pch=data_shapes, bg=data_colours, col="black", xlab="Sepal Length", ylab="Sepal Width")

# or we can modify the histogram
hist(a_nice_table$Sepal.Width, col="indianred", breaks=20)
hist(a_nice_table$Sepal.Width, col="steelblue", breaks=50)
hist(a_nice_table$Sepal.Width, col="forestgreen", breaks=100)

# or add colour to the boxplot
boxplot(Sepal.Length ~ Species, data = a_nice_table, col=c("coral", "midnightblue", "darkgoldenrod1"))

```

Oh, that is better. But wait...how do a person know which species is which? The legend is missing, so you need to add it. A legend is something that you add _on top_ of the plot, it's not included in the `plot()` function (no signs of a legend in the `help(plot)` right?). Let's have a look at `help(legend)`.

```{r}

# scatterplot with colours
plot(x=a_nice_table$Sepal.Length, y=a_nice_table$Sepal.Width, pch=data_shapes, bg=data_colours, col="black", xlab="Sepal Length", ylab="Sepal Width")

# add a legend, and that's it!
legend("bottomright", c("setosa","virginica", "versicolor"), col=colours_i_like, pch=shapes_i_like)

```

# GGPlot2 Plotting

Aside from base R plotting, another very well known method is based on `ggplot2`. `ggplot2` offers many, many different possibilities to obtain any kind of plot. [This](https://r-graph-gallery.com/) website is full of examples, that can help you get started with the plots you want to have.

However, before using `ggplot2`, let's discuss how it works.

The idea behind ggplot2 is _modularity_. If you have ever seen how a `ggplot2` plot is created you might have noticed that usually it takes more than one line. Each line is called a layer, and it define one characteristic of the final plot.

To have a better understanding, create three different plots, using the same structure:

```{r}

# load ggplot2
library("ggplot2")

# load a table from a csv file
a_nice_table <- read.csv("/home/gabriele/work/microbiology/think_like_a_programmer/R_foundations/lessons_files/data/a_nicely_formatted_table.csv", header=T, sep=",")

# plot using dots
ggplot(a_nice_table, aes(x=Sepal.Length, y=Sepal.Width, col=Species)) +
  geom_point()

# same plot, with boxplots
ggplot(a_nice_table, aes(x=Sepal.Length, y=Sepal.Width, fill=Species)) +
  geom_boxplot()

# same plot, with barplot
ggplot(a_nice_table, aes(x=Sepal.Length, fill=Species)) +
  geom_bar()
```

Now, what you just did was to pass `a_nice_table` to `ggplot2` through the `ggplot()` function, then specified some aesthetic parameters. Finally, you added another layer, using the `+`, specifying the type of plot you wanted.

At this point, you can add more specifications such as the colour, for example:

```{r}

# plot using dots
ggplot(a_nice_table, aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point(aes(colour=Species)) +
  scale_colour_manual(values=c("lightcoral", "midnightblue", "saddlebrown"))
```

Now, add some order to the labels, if you like. That's where **factors** become interesting:

```{r}

# load a table from a csv file
a_nice_table <- read.csv("/home/gabriele/work/microbiology/think_like_a_programmer/R_foundations/lessons_files/data/a_nicely_formatted_table.csv", header=T, sep=",")

# set order of levels for the colouring
a_nice_table$Species <- factor(a_nice_table$Species, levels=c("virginica", "setosa", "versicolor"))

# plot using dots
ggplot(a_nice_table, aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point(aes(colour=Species)) +
  scale_colour_manual(values=c("lightcoral", "midnightblue", "saddlebrown"))
```

Finally, change the style using `themes` and `labels` of the plot.

As you can see, a palette with three colours was defined. Then, using the function `names()`, the species were mapped to the colours. This is fundamental for `ggplot2` to know which colour will be used for which species.

```{r}

# load a table from a csv file
a_nice_table <- read.csv("/home/gabriele/work/microbiology/think_like_a_programmer/R_foundations/lessons_files/data/a_nicely_formatted_table.csv", header=T, sep=",")

# set order of levels for the colouring
a_nice_table$Species <- factor(a_nice_table$Species, levels=c("virginica", "setosa", "versicolor"))
levels(a_nice_table$Species) <- c("Virginica", "Setosa", "Versicolor")

# define palette
palette <- c("lightcoral", "midnightblue", "saddlebrown")
# define names of each colour to map species to colour
names(palette) <- levels(a_nice_table$Species)

# plot using dots
ggplot(a_nice_table, aes(x=Sepal.Length, y=Sepal.Width, fill=Species)) +
	geom_point(size=4, stroke=1, color="black", shape=21) +
	scale_fill_manual(values=palette) +
	guides(fill=guide_legend(override.aes=list(shape=21))) +
  xlab("Sepal length") +
  ylab("Sepal width") +
  ggtitle("A nice scatterplot") +
  theme_bw() +
  theme(text=element_text(size=15, face="bold"), legend.position="right")

```

# Solutions to exercises in lesson 04

### If-else

```{r}

# load contacts
contact_list <- read.csv("/home/gabriele/work/microbiology/think_like_a_programmer/R_foundations/lessons_files/data/contacts_list.csv", header=T, sep="\t")

# define two random person names
a_person <- "Silvia"
a_person <- "Elena"

# find a_person in the contact list
person_index <- which(contact_list$name==a_person)

# was someone found?
if (length(person_index) > 0) {
  print(paste(a_person, "phone n. is:", contact_list$phone[person_index], sep=" "))
} else {
  print(paste(a_person, "was not found.", sep=" "))
}

# ifelse() version
ifelse(a_person %in% contact_list$name, paste(a_person, "phone n. is:", contact_list$phone[which(contact_list$name==a_person)], sep=" "), print(paste(a_person, "was not found.", sep=" ")))
```

### For and While

When we discussed how to write a function to compute the average of a set of values, we started doing it for vectors of two, and then we generalised to any vector.

To generalise, we used the function `sum()`.

Can you replicate what the function `sum()` is doing:
 
 - using a `for` loop?
 
```{r}

sum_for <- function(a_vector){
  
  # define variable at 0
  total_sum <- 0
  
  # loop through the vector's values
  for (a_value in a_vector) {
    # sum the current value to zero
    total_sum <- total_sum + a_value
  }
  
  # return the total sum
  return(total_sum)
}

```

 - using a `while` loop?

```{r}

sum_while <- function(a_vector){
  
  # define variable at 0
  total_sum <- 0
  
  # loop through the vector's values
  while (length(a_vector)!=0) {
    # sum the current value to zero
    total_sum <- total_sum + a_vector[length(a_vector)]
    
    # remove the value that was just summed from
    # the vector
    a_vector <- a_vector[-length(a_vector)]
  }
  
  # return the total sum
  return(total_sum)
}

```

Can you use the loop to re-write the `arithmetic_mean()` function?

```{r}

# define the function average for two values
# however by default, a_set_of_values will be equal to
# the vector c(6, 8) and the function, if no input was
# specified, will return 7, which is the average of 6 and 8
arithmetic_mean <- function(a_set_of_values=c(6, 8)) {
  # sum_for the two values and divide by the number of values
  the_sum <- sum_for(a_set_of_values)
  
  # sum_for the two values and divide by the number of values
  the_sum <- sum_while(a_set_of_values)
  
  # compute the number of values in the vector
  the_n_of_vals <- length(a_set_of_values)
  # compute the average
  average <- the_sum/the_n_of_vals
  # return the average value you just computed
  return(average)
}

# run function with no input: default values will be used
arithmetic_mean()
# run function with input: default will not be used
# instead the average of 10 and 3 will be returned
arithmetic_mean(c(10, 3))

```

# Final exercise

Load the data, i.e. `counts_final.csv` and `metadata_final.csv` tables

 1. difficult to visualise, easy to plot: for each phylum plot the abundances (that are stored in the counts table), to compare treatment and controls, in winter and summer, DNA and RNA (these info are stored in the metadata table);
 
 2. medium to easy to visualise if you did n. 1. difficult to plot (probably): visualise the variables vegetation and pH together to see if they are correlated (use a scatterplot, for example). Colour the data points by phylum. Or color them by season, or treatment, or DNA/RNA Or you can colour with a variable and shape with another variable. Or, and this is what I suggest to do, is to create two plots, one for DNA, one for RNA, and to colour by treatment, and shape by season. The phylum can be added as a label. To add labels in a `ggplot2` plot, you may want to check out `geom_text()`.
